Easy as Pi: A Network Coding Raspberry Pi Testbed

### 2016-06-03 raspi_journal.tex

%=================================================================
\documentclass[electronics,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}
\synctex=1
\usepackage{fixme}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{listings}             % Include the listings-package
\usepackage{placeins}
\usepackage{xcolor}
\usepackage{caption}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\definecolor{mygreen}{rgb}{0,0.6,0}
\definecolor{mygray}{rgb}{0.5,0.5,0.5}
\definecolor{mymauve}{rgb}{0.58,0,0.82}
\lstset{ %
  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
  basicstyle=\footnotesize\ttfamily,
  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
  breaklines=true,                 % sets automatic line breaking
  captionpos=b,                    % sets the caption-position to bottom
  %commentstyle=\color{mygreen},    % comment style
  deletekeywords={...},            % if you want to delete keywords from the given language
  %escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
  escapeinside={<@}{@>}
  extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
  frame=single,	                   % adds a frame around the code
  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
  %keywordstyle=\color{blue},       % keyword style
  %language=Bash,                 % the language of the code
  otherkeywords={*,...},           % if you want to add more keywords to the set
  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
  numbersep=5pt,                   % how far the line-numbers are from the code
  numberstyle=\tiny\color{mygray}, % the style that is used for the line-numbers
  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
  showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
  showstringspaces=false,          % underline spaces within strings only
  showtabs=false,                  % show tabs within strings adding particular underscores
  stepnumber=1,                    % the step between two line-numbers. If it's 1, each line will be numbered
  %stringstyle=\color{mymauve},     % string literal style
  tabsize=2,	                   % sets default tabsize to 2 spaces
  title=\lstname,                  % show the filename of files included with \lstinputlisting; also try caption instead of title
  xleftmargin=5mm,
  xrightmargin=5mm,
  columns=flexible
}

>>>>

%=================================================================
\documentclass[electronics,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}
\synctex=1
\usepackage{fixme}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{listings}             % Include the listings-package
\usepackage{placeins}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{textcomp}  % To get the correct quotes in lstlistings (')

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\definecolor{mygreen}{rgb}{0,0.6,0}
\definecolor{mygray}{rgb}{0.5,0.5,0.5}
\definecolor{mymauve}{rgb}{0.58,0,0.82}
\lstset{ %
  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
  basicstyle=\footnotesize\ttfamily,
  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
  breaklines=true,                 % sets automatic line breaking
  captionpos=b,                    % sets the caption-position to bottom
  %commentstyle=\color{mygreen},    % comment style
  deletekeywords={...},            % if you want to delete keywords from the given language
  %escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
  escapeinside={<@}{@>}
  extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
  frame=single,	                   % adds a frame around the code
  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
  %keywordstyle=\color{blue},       % keyword style
  %language=Bash,                 % the language of the code
  otherkeywords={*,...},           % if you want to add more keywords to the set
  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
  numbersep=5pt,                   % how far the line-numbers are from the code
  numberstyle=\tiny\color{mygray}, % the style that is used for the line-numbers
  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
  showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
  showstringspaces=false,          % underline spaces within strings only
  showtabs=false,                  % show tabs within strings adding particular underscores
  stepnumber=1,                    % the step between two line-numbers. If it's 1, each line will be numbered
  %stringstyle=\color{mymauve},     % string literal style
  tabsize=2,	                   % sets default tabsize to 2 spaces
  title=\lstname,                  % show the filename of files included with \lstinputlisting; also try caption instead of title
  xleftmargin=5mm,
  xrightmargin=5mm,
  columns=flexible,
  upquote=true
}

### 2016-06-03 testbed.tex
more testbed

\begin{lstlisting}[frame=single]  % Start your code-block
IMAGE="raspbian_lite-2016-05-31/2016-05-27-raspbian-jessie-lite"
WORKDIR="${HOME}/Raspbian"
cd ${WORKDIR}
wget http://downloads.raspberrypi.org/raspbian_lite/images/${IMAGE}.zip
unzip ${IMAGE}
\end{lstlisting}
\FloatBarrier

We have installed the official Raspbian linux \ac{OS} on all the devices.

\subsection{Installation and Updating}
\subsection{Kodo cross-compilation: From your PC to the Raspberry Pi}

>>>>

Open a terminal, and then...

hashmark (\#) means to perform operation as root, dollar (\$) means to perform operation as user

% Define variable]
\begin{lstlisting}[] 
$ export IMAGE="2016-05-27-raspbian-jessie-lite"
$ export WORKDIR="${HOME}/Raspbian"
\end{lstlisting}
\FloatBarrier

% Download and unpack image
\begin{lstlisting}[] 
$ mkdir -p ${WORKDIR}
$ cd ${WORKDIR}
$ wget http://downloads.raspberrypi.org/raspbian_lite/images/raspbian_lite-2016-05-31/${IMAGE}.zip
$ unzip ${IMAGE}.zip
\end{lstlisting}
\FloatBarrier

% Check out the image   % The dollar hack was to fix vim syntax
\begin{lstlisting}[literate={DOLLAR}{\$}1]
DOLLAR fdisk -lu ${IMAGE}.img
Disk 2016-05-27-raspbian-jessie-lite.img: 1.3 GiB, 1387266048 bytes, 2709504 sectors
Units: sectors of 1 * 512 = 512 bytes
Sector size (logical/physical): 512 bytes / 512 bytes
I/O size (minimum/optimal): 512 bytes / 512 bytes
Disklabel type: dos
Disk identifier: 0x6fcf21f3

Device                               Boot  Start     End Sectors  Size Id Type
2016-05-27-raspbian-jessie-lite.img1        8192  137215  129024   63M  c W95 FAT32 (LBA)
2016-05-27-raspbian-jessie-lite.img2      137216 2709503 2572288  1.2G 83 Linux
\end{lstlisting}
\FloatBarrier

We see that the boot and root partitions starts at sector 8192 and 137216 respectivly.
We also notice that each sector is 512 bytes.

On newer systems, we can mount the device easier using "losetup --show -f -P IMAGE"

Mount root partition:

\begin{lstlisting}[]
$ export ROOTDIR="${WORKDIR}/root"
$ mkdir -p ${ROOTDIR}
# mount -o loop,offset=$((137216*512)) ${IMAGE}.img ${ROOTDIR}
\end{lstlisting}
\FloatBarrier

Mount boot partition:
\begin{lstlisting}[]
# mount -o loop,offset=$((8192*512)) ${IMAGE}.img ${ROOTDIR}/boot
\end{lstlisting}
\FloatBarrier

We can now change all the files in the disk image.

\subsection{Configuring OS files and scripts}

% MAC and Hostname file
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/.rasp\_config/nodes.csv}@>
<@\textcolor{gray}{
---------------------------------------------------------------}
\Reactivatenumber @>
# Ethernet MAC    Hostname
b8:27:eb:5b:da:20 rasp_00
b8:27:eb:7b:c3:91 rasp_01
b8:27:eb:54:9c:64 rasp_02
b8:27:eb:95:bd:11 rasp_03
\end{lstlisting}
\FloatBarrier

% Set hostname
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/.rasp\_config/set\_hostname}@>
<@\textcolor{gray}{
---------------------------------------------------------------}
\Reactivatenumber @>
#!/usr/bin/env bash
mac=$(cat /sys/class/net/eth0/address)
hostname=$(grep $mac nodes.csv | cut -f2 -d' ')

# Assign hostname found in nodes.csv
if [ ! -z $hostname ]; then
    echo $hostname > /etc/hostname
    hostname $hostname
fi
\end{lstlisting}
\FloatBarVrier

% Get the files
\begin{lstlisting}[]
wget http://kom.aau.dk/project/TuneSCode/rasp_config.zip
unzip rasp_config.zip -d ${ROOTDIR}/home/pi/
\end{lstlisting}
\FloatBarrier

Call set\_hostname script at startup. We insert a line of code to call script in rc.local just after the initial comments (i.e. lines starting with \#).
\begin{lstlisting}[]
line_number=$(egrep -n -m1 "(^[^#])|(^$)" ${ROOTDIR}/etc/rc.local | cut -d: -f1)
sed -i "${line_number}c bash /home/pi/.rasp_config/set_hostname" ${ROOTDIR}/etc/rc.local
\end{lstlisting}
\FloatBarrier
%$ line="bash /home/pi/.rasp_config/set_hostname"
%# awk -v text="$line" '!/^#/ && !p {print text; p=1} 1' ${ROOTDIR}/etc/rc.local > <@ \Suppressnumber @>
%    ${ROOTDIR}/etc/rc.local <@ \Reactivatenumber @>

All other files could be updated.

\subsection{Installation and Updating}

It may be desired to pre-install some programs in the image before it is
written to all the memory cards that goes into the Raspberry pis.
This can be done using QEMU Chroot (change root).
\url{https://wiki.archlinux.org/index.php/Raspberry_Pi}

\fxnote{Lets use RPi for Raspberry Pis throughout the paper!}
\fxnote{Which packages is required for this on debian/ubuntu?}

% CHROOT to OS image
\begin{lstlisting}[]
# apt-get install binfmt-support qemu qemu-user-static
# update-binfmts --importdir /var/lib/binfmts/ --import
# update-binfmts --display qemu-arm
# update-binfmts --enable qemu-arm
\end{lstlisting}
\FloatBarrier

% CHROOT to OS image
\begin{lstlisting}[]
# cd $ROOTDIR
# cp /etc/resolv.conf etc/resolv.conf
# cp /usr/bin/qemu-arm-static usr/bin
# mount -t proc proc proc/
# mount --rbind /sys sys/
# mount --rbind /dev dev/
# mount --rbind /run run/
# chroot ${ROOTDIR} /usr/bin/qemu-arm-static /bin/bash

ALTERNATIVE:
# proot -q qemu-arm-static -S ${ROOTDIR}
\end{lstlisting}
\FloatBarrier

% Update system
\begin{lstlisting}[]
# apt-get update
# apt-get upgrade
\end{lstlisting}
\FloatBarrier

Lets install some programs
% Install packages
\begin{lstlisting}[]
# apt-get install vim git
\end{lstlisting}
\FloatBarrier

\subsection{Installation an overlaying filesystem}

We use the code provided by

Assumuing that we are still chrooted to the RPi image and the boot partition is mounted to /boot

% Get files
\begin{lstlisting}[]
# OVERLAYROOTDIR="/tmp/overlayroot"
# git clone https://github.com/chesty/overlayroot.git $OVERLAYROOTDIR
\end{lstlisting}
\FloatBarrier

% Install required packages
\begin{lstlisting}[]
# apt-get install busybox
\end{lstlisting}
\FloatBarrier

% Tell uboot to load initramfs
\begin{lstlisting}[]
# echo "initramfs init.gz" > /boot/config.txt
\end{lstlisting}
\FloatBarrier

% Setup initramfs
\begin{lstlisting}[]
# echo "overlay" > /etc/initramfs-tools/modules
# cp ${OVERLAYROOTDIR}/hooks-overlay /etc/initramfs-tools/hooks/
# cp ${OVERLAYROOTDIR}/init-bottom-overlay /etc/initramfs-tools/scripts/init-bottom/
\end{lstlisting}
\FloatBarrier

% Create an initramfs
\begin{lstlisting}[]
# mkinitramfs -o /boot/init.gz
\end{lstlisting}
\FloatBarrier

\subsection{Kodo cross-compilation: From your PC to the Raspberry Pi}
